---
title: "Stats 306: Lecture 2"
subtitle: "Basic R,  Tables, Vectors, and ggplot"
author: "Jayashree Ravi"
citation: Content adapted from slides by Dr. Mark Fredrickson
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
options(readr.show_progress = T)
```

## Review - Expressions and Statements

When a program executes, you typically create temporary variables in the RAM of your computer. The values that these variables keep at any point of time during your R program execution is called its **state**.

An **expression** is R code that, when run, provides a value (do not change state):

```{r eval = F}
3 + 4
```

Commands that change the **state** of the program are called **statements**, such as assignment:

```{r}
a <- 3 + 4
```



**Variables and Assignment**

A **variable** is a box to store data. We put the results of expressions into variables using **assignment**.

```{r}
a <- 7
b <- 12
c <- a + b
print(c)
```

Rules for variable names: start with a letter then letters, numbers, `.`, `_`. 




## The REPL and Chunks

When you **knit** an RMarkdown document, RStudio **starts a new process**.

Every time it encounters a **R chunk** it **evaluates** the R code inside and inserts the output into the document (default; can be changed). Like a behind the scenes REPL.


**Order is important**

Because **statements** change the context, order of statements is very important.

```{r warning=FALSE}
a <- 7
a <- 2
print(a)  
```

**Working with code chunks, debugging hint**

If you run into a problem (for e.g., a weird behavior that you can't explain logically), use the `Session -> Restart R` to clean up, and the use "Run all chunks above". Then try the chunk again. 

## What can go in variables?

R has several different **data types**. Some common types:

* **Numeric**: `7`, `3.22`, `9.1e2`
* **Strings**: `"Hello"` or `'Hello'` (also called "character vectors")
* **Logicals**: `TRUE` and `FALSE`
* We'll see more data types going forward
* These are also referred to as the **class** of the data.

## Why are variables useful?

* Avoid repeatedly typing in the same thing
* Save the results of computations
* Avoid recomputing same values
* Separate what we are computing from any particular input (**abstraction**)

**Exercise: Using variables**

Find the difference between 193 and 7. What is this value divided by two? What is this value divided by three? (Only compute the difference once.)
```{r using-variables, exercise = TRUE, exercise.lines = 5}

```

## Style point: descriptive variable names

* So far we have used names like `a` and `b`.
* For large programs, having variable names that describe what they hold is much better
* Compare `a` to `number_of_seconds_in_a_day`
* In RStudio, it's just as easy to type with `TAB` completion
* On HW assignments, we will awarding for good *style* (readable code)
* You are encouraged to follow the coding style suggested by tidyverse: https://style.tidyverse.org/

## Functions 
A key tool in any programming language is a **function**.

* R functions behave (mostly) like mathematical functions $f(x)$.
* They take **arguments** and return **values**
* Encapsulate code (mini-programs)
* Make it easy to repeat operations

**Syntax**

```{r, eval = FALSE}
return_value <- f(argument1, argument2, named_argument = additional_data, ...)
```

* Like math functions, we use the syntax `f()` to run the function
* All **arguments** are **passed** to the function inside of the `()` separated by commas
* Arguments can be passed in **positional** order or
* Passed by **name**
* Some arguments are **optional** when they have a default value
* `...` means "any number of additional arguments"
* `return_value` is the output of the function


## Defining functions

```{r, results='hide'}
mult <- function(x, y) { 
  print(paste('x = ' , x))
  x * y 
}
```

The `mult` function takes in two arguments and multiplies them

```{r}
mult(2, 10)
mult(y = 5, x = 3)
```
There are many built-in functions in R that you can simply use.

[Glimpse at R Vocabulary](http://adv-r.had.co.nz/Vocabulary.html)


## Function evaluation

In R, when we call a function, first the **arguments** are evaluated, then the function is called.

```{r}
mult(2, mult(2,3))
```


## Avoiding single use variables

Sometimes we only need to use something once, so taking the time to come up with a variable name is unnecessary. 

Instead we can use **pipes** to pass the output of one function to the **first argument** of a second function.

```{r, eval = FALSE}
value <- f()
g(value)

# pipes
f() |> g()
```

```{r}
mult(2.5, 3) |> round()
```

In older R programs, you also see another notation for pipe `%>%`.
With this notation the same expression can be 

```{r}
mult(2.5, 3) %>% round
```

Note: We did not have to put the pair of parenthesis for this notation. If you do put, it doesn't hurt


## R's assignment operators

R has two ways to store data in variables:

```{r}
variable_arrow <- 7
variable_equal = 10
variable_arrow + variable_equal
```

Arrow can also be reversed

```{r}
7 -> variable_reversed
pi * variable_reversed
pi |> sqrt() -> sqrt_pi
sqrt_pi
```

There is a subtle difference between using `=` and `->`.  For example, the below function does not work if you use arrow. So for any function call key/value arguments, you have to use `=`


```{r}
x = c(10)
try_me <- function(y = x){
  print(y)
}
try_me(x)

```

## Some special values

R has special data values

* `NA` means "missing". Including a `NA` value in a calculation will usually make the result also `NA`.
* `NULL` means "not set yet" and is often used for optional arguments to functions
* `NaN` means something was divided by zero or other mathematical impossibility.

What will happen with this code?

```{r na-prop}
values = c(1, 2, NA)
mean(values)
```

Instead add the keyword argument `na.rm = T` to instruct the function to first remove all the NA values and then compute the mean

```{r}
mean(values, na.rm = T)
```
Let us try a few more expressions with NA

```{r nan-tryouts, exercise = TRUE, exercise.lines = 5}

```


## Lists and Vectors

* A **vector** is an ordered collection of the same type of data. Created using the `c` function (concatenate).
* A **list** is an ordered collection of (potentially) different types of data. Created using the `list` function.
* Both can have **names** to indicate individual items.

**Vector example**

```{r}
phone_digits <- c(8, 6, 7, 4, 3, 0, 9)
phone_plus_area_code <- c(7, 3, 4, phone_digits)
print(phone_plus_area_code)
length(phone_plus_area_code)
sum(phone_plus_area_code)
str_flatten(phone_plus_area_code)
```

We will learn more on *str* functions when we learn Strings.

**List example**

Lists are heterogeneous while vectors are homogeneous

```{r}
mixed_type_list <- list(3, 4, phone_digits)
mixed_type_list
list("umich", 1, T)
c("umich", 1, T)
```



## Vectors, indexing

The `runif` function will create randomly generated values in the range (0, 1).

```{r}
random_u01 <- runif(10) # create 10 values
print(random_u01)
random_u01[1] # first item
random_u01[10] # last item
random_u01[-1] # all but first
random_u01[1:3] # first three
random_u01[c(1, 3, 7, 1)]
```



## Relational Operators

```{r}
x <- 1
y <- 2

x < y
x > y
x <= y
x >= y
x == y
x != y

!(x == y) # negation
```

## Vectors, subsetting using relational operators

```{r}
random_u01
less_than_0.5 <- random_u01 < 0.5
print(less_than_0.5)
random_u01[less_than_0.5]
keep(random_u01, random_u01 < 0.5)
random_u01[!less_than_0.5]
```


## Named collections

Vectors can have names (`c(first = 1, second = 22, third = 333)`), but names for lists are more common.
```{r}
named_list <- list(a_string = "A string", a_vector = 1:4)
named_list[["a_string"]] # note double brackets
named_list$a_vector
```


## Tables in R

In data analytics we often work with Tabular data where:

>* columns are variables
>* rows are observations

A tabular data is presented using a `tibble` or `data.frame`.
When we construct new tabular data, we will use `tibble` as that is a better data structure than `data.frame`. However many existing libraries use `data.frame`.
We will treat them interchangeably for the most part.

You construct a simple tibble with two columns named `a` and `b` as shown below

```{r}
tibble(a = c(1:10),
       b = c(11:20))
```

There are many tabular datasets readily available when we load `tidyverse` among them `mpg` is one of them. There are many functions you can use on the tibble and here we will see some of them

```{r}
glimpse(mpg)
class(mpg)
dim(mpg)
colnames(mpg)
head(mpg)
```

## Accessing individual columns

Recall, the `$` operator lets us get items from lists:

```{r}
mpg$hwy |> mean() # equiv to mean(mpg$hwy)
```

## Other functions for basic descriptive statistics

```{r}
sum(mpg$hwy)
var(mpg$hwy)
sqrt(var(mpg$hwy)) # standard deviation
median(mpg$hwy)
IQR(mpg$hwy)
quantile(mpg$hwy, c(0.25, 0.75))
```

All at once

```{r}
summary(mpg)
```

## Dealing with missingness

To delete the missing values we have some options. Let us look into the `storms` data; another tabular data provided by NOAA Atlantic hurricane database representing the positions and attributes of storms from 1975-2022. 

```{r}
storms$category |> mean()
storms$category |> mean(na.rm = T)
na.omit(storms$category) |> mean()
storms$category[!is.na(storms$category)] |> mean()
```

## Plotting: Visual representation of data

**Graphing** or **plotting** is the representation of data in a visual form, typically on a 2D plane.

Humans have a wonderful ability to process visual stimuli quickly.

Great for displaying large data sets that might be difficult to describe.

**But Plotting ain't easy**

* What exactly can we put in a plot?
* How to connect the plot to data?
* What are our goals for the plot?
* What additional information is necessary to understand the plot?

## The Semiology of Graphics

In 1967, **Jacques Bertin** published the *Semiology of Graphics* to describe common elements of plots and what they could achieve.

Bertin described two ways of thinking about plots:

1. Visual ("retinal") variables: connections between objects in the plot and underlying data
2. Relationship what types of relationships can the visual variables express

## Visual variables/retinal variables:

* Position/planar dimensions
* Size (small to large)
* Value (light to dark)
* Texture (pattern)
* Color (discrete or continuously varying)
* Orientation (angle/rotation)
* Shape (usually distinct shapes)



[Visual Variables](https://en.wikipedia.org/wiki/Visual_variable)

## The Grammar of Graphics and `ggplot`

A follow up to Bertin's work was *The Grammar of Graphics* by Leland Wilkinson.
This book described a programming language for graphics based on ideas in
Bertin's system.

The GoG was implemented for R in `ggplot` (and later replaced by `ggplot2`). 

## The main components of a ggplot graph

* The graph object itself (creating using `ggplot(data)`)
* A set of *aesthetic* mappings (connecting data to visual variables)
* Layers: collections of geometric elements (`geom_*()`) and statistical transformations (`stat_*()`)
* Scales: information on the range or composition of variables
* Coordinate systems: how the data are arranged spatially
* Facet: breaking a single plot into many, similar plots by grouping the data into multiple facets
* Theme: all the other color and printing aspects of the plot

## Creating a ggplot

Start use the `ggplot` function to start our plot
```{r}
efficiency <- ggplot(data = mpg)
efficiency # for now, blank
```



**What can we plot?** Let us take a glimpse of the dataset
```{r}
glimpse(mpg)
```

"Aesthetic" mappings connect columns to visual variables
```{r}
efficiency <- ggplot(data = mpg,
                     aes(x = displ, y = hwy, color = cyl))

efficiency
```
displ: engine displacement is the measure of the total volume of all the cylinders in an internal combustion engine.
hwy: highway miles/gallon
cyl: number of cylinders


But the chart is still blank? Well we need to add the geometric layer to define how to display the visual elements

## Geometries: objects on the plot

We will use a **geometry function** (have the form `geom_TYPE()`).

```{r}
efficiency + geom_point()
```

**Alternative forms**

```{r eval = F}

ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy, color = cyl))

mpg |> ggplot() + geom_point(aes(x = displ, y = hwy, color = cyl))
```

## Exercise: Great Lakes water levels

Here's a data set from the US Army Corps of Engineers showing the height of the Great Lakes (with Michigan and Huron being one body) above sea level, aggregate by year.

```{r}
glwl <- read_csv("data/GLHYD_data_english_cleaned.csv.gz") |>
  group_by(body, year) |> summarize(ht = median(height_ft)) |> ungroup()
glwl
```

Use this data set to make a scatter plot of the height of each lake in each year, colored by which `body` the reading is from.

```{r glwl-plot-setup}

glwl <- read_csv("data/GLHYD_data_english_cleaned.csv.gz") |>
  group_by(body, year) |> summarize(ht = median(height_ft)) |> ungroup()
```

```{r glwl-plot, exercise = TRUE}

```

Improve your plot by changing it to a `geom_line()`

## What aesthetic mappings can I use?


**Trying out some other mappings**

```{r}
ggplot(data = mpg,
       aes(x = displ, y = hwy, size = cyl, color = class)) + 
  geom_point()
```

Visit the docs to learn more: [geom_point](https://ggplot2.tidyverse.org/reference/geom_point.html)

## Using expressions

We can also use expressions involving columns.

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, shape = year > 2000)) + 
  geom_point()
```

## Overriding parameters for all points

We can pass in constants that apply to all points (size and transparency):

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, shape = year > 2000)) + 
  geom_point(size = 5)
```

## Jitter: useful noise

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) + 
  geom_point(position = "jitter")
```


## List of geometries

[More ggplot documentation](https://ggplot2.tidyverse.org/reference/index.html#geoms).


## General observations

* Generally limit plots to having 3 or 4 distinct visual variables
* Typically only one geometry type, but there are some great counter examples
* Almost everything can be tweaked in ggplot, finding it is the tricky part
* The [R Graph Gallery](https://r-graph-gallery.com/) is a great source of inspiration and instruction

## `homework/example-homework.Rmd`

Using an example homework file, we'll answer a few questions and prepare our HW for submission.

After running all code, it turns the markdown into a web page (`example-homework.Rmd` => `example-homework.html`)