---
title: "Stats 306: Lecture 1"
subtitle: "Basic R and RMarkdown"
author: "Jayashree Ravi"
citation: Content adapted from slides by Dr. Mark Fredrickson
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
```



## Expressions and Statements

When a program executes, you typically create temporary variables in the RAM of your computer. The values that these variables keep at any point of time during your R program execution is called its **state**.

An **expression** is R code that, when run, provides a value (do not change state):

```{r eval = F}
3 + 4
```

Commands that change the **state** of the program are called **statements**, such as assignment:

```{r}
a <- 3 + 4
```

## Variables and Assignment

A **variable** is a box to store data. We put the results of expressions into variables using **assignment**.

```{r}
a <- 7
b <- 12
c <- a + b
print(c)
```

Rules for variable names: start with a letter then letters, numbers, `.`, `_`. 

## REPL and State

Suppose you entered the following at your R console:

```{r repl-state-setup}
a <- 3 + 1
a <- 10 + 2
```
What would happen if you did the following?
```{r repl-state, exercise = TRUE}
print(a)
```

## The REPL and Chunks

When you **knit** an RMarkdown document, RStudio **starts a new process**.

Every time it encounters a **R chunk** it **evaluates** the R code inside and inserts the output into the document (default; can be changed). Like a behind the scenes REPL.


## Order is important

Because **statements** change the context, order of statements is very important. Compare
```{r}
a <- 7
print(a)
a <- 2
```

```{r warning=FALSE}
a <- 7
a <- 2
print(a)  
```

## Working with code chunks, debugging hint

If you run into a problem, use the `Session -> Restart R` to clean up, and the use "Run all chunks above". Then try the chunk again. You could also use the keyboard shortcuts suggested next to the button too.

## What can go in variables?

R has several different **data types**. Some common types:

* **Numeric**: `7`, `3.22`, `9.1e2`
* **Strings**: `"Hello"` or `'Hello'` (also called "character vectors")
* **Logicals**: `TRUE` and `FALSE`
* We'll see more data types going forward
* These are also referred to as the **class** of the data.

## Why are variables useful?

* Avoid repeatedly typing in the same thing
* Save the results of computations
* Avoid recomputing same values
* Separate what we are computing from any particular input (**abstraction**)

## Exercise: Using variables

Find the difference between 193 and 7. What is this value divided by two? What is this value divided by three? (Only compute the difference once.)
```{r using-variables, exercise = TRUE, exercise.lines = 5}

```

## Style point: descriptive variable names

* So far we have used names like `a` and `b`.
* For large programs, having variable names that describe what they hold is much better
* Compare `a` to `number_of_seconds_in_a_day`
* In RStudio, it's just as easy to type with `TAB` completion
* On HW assignments, we will be awarding points for *style* (readable code)

## Functions 
A key tool in any programming language is a **function**.

* R functions behave (mostly) like mathematical functions $f(x)$.
* They take **arguments** and return **values**
* Encapsulate code (mini-programs)
* Make it easy to repeat operations

## Functions in R

Once we have data stored in a variable, one of the main ways to use it is to **call a function.**

```{r, eval = FALSE}
return_value <- function_name(argument1, argument2, named_argument = additional_data, ...)
```

* Like math functions, we use the syntax `f()` to run the function
* All **arguments** are **passed** to the function inside of the `()` separated by commas
* Arguments can be passed in **positional** order or
* Passed by **name**
* Some arguments are **optional** when they have a default value
* `...` means "any number of additional arguments"
* `return_value` is the output of the function


## Converting from degrees to radians

Recall two ways to measure an angle:

>* Degrees: 0 to 360
>* Radians: 0 to 2$\pi$

Exercise: compute the radian equivalent of 45 degrees (use a good variable name!). Then use the `sin(your_variable_name)` function to compute the sine of 45 degrees. (Hint: there is a built in variable named `pi`.)

```{r sin45, exercise = TRUE}

```

## Defining functions

```{r, results='hide'}
mult <- function(x, y) { x * y }
div <- function(x, y) { x / y }
```

Suppose we had `mult(x, y)` and `div(x, y)`.

We could have written the previous calculation as 

```{r}
mult(2, mult(pi, div(45, 360)))
```

[Glimpse at R Vocabulary](http://adv-r.had.co.nz/Vocabulary.html)

## `homework/example-homework.Rmd`

Using an example homework file, we'll answer a few questions and prepare our HW for submission.

After running all code, it turns the markdown into a web page (`example-homework.Rmd` => `example-homework.html`)


## Function evaluation

In R, when we call a function, first the **arguments** are evaluated, then the function is called.

```{r}
sin(mult(2, mult(pi, div(45, 360))))
```
Equivalent to:
```{r}
value1 <- div(45, 360)
value2 <- mult(pi, value1)
value3 <- mult(2, value2)
sin(value3)
```

## Avoiding single use variables

Sometimes we only need to use something once, so taking the time to come up with a variable name is unnecessary. 

Instead we can use **pipes** to pass the output of one function to the **first argument** of a second function.

```{r, eval = FALSE}
value <- f()
g(value)

# pipes
f() |> g()
```

Chaining pipes:
```{r}
div(45, 360) |>  
  mult(pi) |>
  mult(2) |>
  sin()
```

## Functions as variables

In R, **functions are variables** which means that we can use them like we would any other variable.

```{r}
print(10)
print(mean)
```



## The `curve` function

The `curve` function is a useful way to graph out a function:
```{r, height = 4, width = 7}
curve(sin)
```


## Changing the bounds of the plot

`curve` can take more than just one argument,

```{r, height = 3, width = 5}
curve(sin, 0, 2 * pi)
```

## Optional and named arguments

R allows you to use the name of arguments to pass them in any order:

```{r}
curve(to = 2 * pi, expr = sin, from = 0)
```

By *default* `from = 0` and `to = 1`


## R's assignment operators

R has two ways to store data in variables:

```{r}
variable_arrow <- 7
variable_equal = 10
variable_arrow + variable_equal
```

Arrow can also be reversed

```{r}
7 -> variable_reversed
pi * variable_reversed
pi |> sqrt() -> sqrt_pi
sqrt_pi
```

There is a subtle difference between using `=` and `->`.  For example, the below function does not work if you use arrow. So for any function call key/value arguments, you have to use `=`


```{r}
x = c(10)
try_me <- function(y = x){
  print(y)
}
try_me(x)

```

## Some special values

In addition to `pi`, R has some other built in variables (see help page for `Constants`).

Also several special data values

* `NA` means "missing". Including a `NA` value in a calculation will usually make the result also `NA`.
* `NULL` means "not set yet" and is often used for optional arguments to functions
* `NaN` means something was divided by zero or other mathematical impossibility.

What will happen with this code?

```{r na-prop, exercise = TRUE}
angle_degrees <- NA
angle_radians <- 2 * pi * angle_degrees / 360
sin(angle_radians)
```

## Summary

* RMarkdown documents: text and code chunks
* R and the REPL
* Variables: boxes to hold data
* Functions: little programs that operate on data

