---
title: "Stats 306: Lecture 11"
subtitle: "Wrapping up EDA, Wrangling"
output:
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this

```


## Review

* Continued our formalization of EDA:
  * What kind of variation for the measurements in in my sample?
  * What kinds of covariation among measurements (relationships)?
* ECDFs, histograms, density plots
* Importance of bin width (histogram) and bandwidth (density)
* Plots: typical values, clusters (how many, narrow/wide), where is variation?
* Missing data (`NA`): dropping, imputing, modeling
* Covariation

## Diamond data

```{r}
diamonds |> head()
```

```{r}
diamonds |> glimpse()
```

## Histogram

```{r}
diamonds |> ggplot(aes(x = price)) + geom_histogram()
```

## Histrogram for all numerical values

```{r}
diamonds |>
  select(where(is.numeric)) |>
  gather() |>
  ggplot(aes(x = value, fill = key)) +
  geom_histogram(binwidth = 0.5, position = "identity", alpha = 0.7) +
  facet_wrap(~key, scales = "free") +
  labs(title = "Histograms of Numerical Variables",
       x = "Value",
       y = "Frequency",
       fill = "Variable") 
```


```{r}
diamonds |>
  select(where(is.numeric)) |>
  gather()
```

##  Conditional Distribution

```{r}
diamonds |> ggplot(aes(x = cut, y = price)) + geom_violin()
```

## Two Quantitative Joint Distribution


```{r}
diamonds |> ggplot( aes(x = carat, y = price, color = cut)) +
    geom_point(alpha = 0.5) + 
    labs(title = 'Joint Distribution of Quantitative Variables',
         x = "Carat",
         y = "Price")
```

## Locally weighted least squares (loess) 

```{r}
diamonds |> ggplot( aes(x = carat, y = price)) +
    geom_point(alpha = 0.5) + stat_smooth() +
    labs(title = 'Joint Distribution with Smooth curve',
         x = "Carat",
         y = "Price")
```
Stratify by cut

```{r}
diamonds |> ggplot( aes(x = carat, y = price, color = cut)) +
    geom_point(alpha = 0.5) + stat_smooth() +
    labs(title = 'Joint Distribution with Stratification',
         x = "Carat",
         y = "Price")
```


## Pair plot

```{r}
diamonds |>
  select(where(is.numeric)) |>
  pairs(col = diamonds$cut)
```

## Exercise - Titanic dataset

Use the titanic data provided below and perform EDA.

```{r}
install.packages("titanic")
library(titanic)
titanic_train |> head()
```


## Next time

>* R for Data Science: 7/17