---
title: "Stats 306: Lecture 12"
subtitle: "Loading Data, Dates"
author: "Jayashree Ravi"
output:
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this
library(nycflights13) # install.packages("nycflights13")
set.seed(2939394)
```


## Data Files

How do we share data?

* Plain text formats - humans can also read!
* Binary formats - only computers can read
* Database applications - SQL/NoSQL languages to CRUD (Create, Read, Update, Delete); data mostly stored in binary format

## Plain text delimited formats

* **Plain text** files are the easiest format to share across platforms, software.
* Somewhat less efficient for file size.
* We often **compress** the files using zip (`.zip`) or (`.gz`), but still basically plain text
* Organizing tabular data:
  * Each **line** of the file is a row of a table
  * Separate columns using a **delimiter** (if csv - comma; if tsv - tab etc.)

  
## Example, comma separated values (CSV)

```
-82.9,42.4,3043540,1703280231,Gratiot Ave & Mapleridge St,"ACCIDENT, HIT & RUN",OUIL,5422,54001,"ACCIDENT, HIT & RUN",2017/03/28 22:00:00+00
```

>* Columns separated by `,`
>* Numeric values interpreted as given
>* Both unescaped (`Gratiot Ave & Mapleridge St`) and escaped (`"ACCIDENT, HIT & RUN"`) strings.
>* Other kinds of data (e.g. date-times like `2017/03/28 22:00:00+00`) just represented as strings.

## Header rows

Most files will include a **header row** that gives column names, but not all!

Some files also have instructional text at the top which needs to be trimmed before loading.

## Loading delimited files

* R has a set of built in functions `read.EXTENSION` like `read.csv` and `read.table` (any white space as separator, "tab" is common).
* The built in functions return `data.frames`
* Tidyverse (in the )`readr` package) has `read_*` instead. Returns `tibbles` (has a few other features: faster, progress bars for big files, less data transformation initially)

## Example

(White space not important)
```{r}
ex_csv <-'
column1,column2,column3
7.1,sometext,01/02/2022
,"has,comma",07/15/2022
'
read_csv(ex_csv)
```

## Exercise

Use the `delim` argument to [`read_delim`](https://readr.tidyverse.org/reference/read_delim.html) to read this data into a table:
```{r doubledash, exercise = TRUE}
d <- '
col1--col2
7--8
--9
10--11
'
```

```{r doubledash-solution}
d <- '
col1--col2
7--8
--9
10--11
'

read_delim(d, delim = '--')
```

## Common issues

* Extra stuff at top of file, use `skip = NUMBER` to skip lines
* We assume column names by default, can indicate no names with `col_names = FALSE` or give a vector of column names
* Missing values: usually cells without values, but you can override with `na = "."`

```{r}
read_csv(
"# First row to skip
// Second row to skip
% Third row to skip
1, 0 , 3
4, 0, 6
7, 8, 0
", skip = 3, na = c('0'), col_names = c("a", "b", "c"))
```

## Skipping comment lines

Some CSVs will come with comments, typically in the form of lines prefaced by #. You can also skip comments line by specifying a comment character.

```{r}
read_csv("
# First comment line
a, b, c
# This separates the header from the data
1, 2, 3
4, 5, 6
# Another comment line
", comment = '#')
```

## Selecting column types

The `col_types` argument allows setting type of some or all columns. 
```{r}
ex_csv

read_csv(ex_csv, col_types = cols(column3 = col_date("%m/%d/%Y")))
```

`col_skip` allows not importing column.

## Exercise

Read in the following table using `read_csv`. Do not read the first row as header.  Make any "NONE" and 'EMPTY' values be missing in R.

```{r settingcols, exercise = TRUE}
csv <- '
1,hello, NONE, EMPTY
0,goodbye,-3, 100
1,NONE,1.111111, EMPTY
'
```
>* [`read_csv`](https://readr.tidyverse.org/reference/read_delim.html)
>* [`cols`](https://readr.tidyverse.org/reference/cols.html)


```{r settingcols-solution}
csv <- '
1,hello, NONE, EMPTY
0,goodbye,-3, 100
1,NONE,1.111111, EMPTY
'
read_csv(csv, na = c('NONE', 'EMPTY'), col_names = F)
         

```
## Saving data

There are complementary `write_csv` and `write_delim` functions.

R has a proprietary format called "RDS". You can use `write_rds` to write a single table or the `save` function to write more than just one variable to a file. Useful for processing data in a file and then saving it for later use.

## Dates

One of the most common data-as-strings we need to deal with are **dates**, **times**, and **date-times**. 

There are lots of different formats, timezones, leapyears to deal with. The `lubridate` package helps:

```{r}
library(lubridate)
```

## Dates and date-times

```{r}
today()
now()
today() |> class()
now() |> class()
```

The UTC stands for "Coordinated Universal Time" (was Greenwich Mean Time)

We've already seen examples of date-times in the flights tibble:
```{r}
select(flights, time_hour) |> print()
```

You should tend to favor working with dates over date-times if possible. The latter are more complicated because of the need to handle time zones.

## The `mdy`, `ymd`, and `dmy` functions

Converting strings in different formats, to dates and times, is easy with these convenience functions:

```{r}
# Only date
mdy("10/01/2022") # Oct. 1
ymd("2022-10-01")
dmy("01.10.2022")
dmy("31-Jan-2017")
dmy("31 1 2017")

# Date and time
ymd_hms("2017-01-31 20:11:59")
mdy_hms("10/01/2022 12:01:37")
mdy_hm("01/31/2017 08:01")

```
`_hms` version will parse `HOURS:MINUTES:SECONDS` while `_hm` will just be `HOURS:MINTUES`. AM/PM can be included, otherwise 24 clock assumed. 

## Setting Timezone

We can select timezone using the nearest major city:
```{r}
mdy_hms("10/01/2022 12:01:37", tz = "America/Detroit")
mdy_hms("12/01/2022 12:01:37", tz = "America/Detroit")
```

## Exercise

Parse the following into dates or date times:
```{r parsed, exercise = TRUE}
dt1 <- "31/10/89"
dt2 <- c("2022-07-01 3:10am",
         "1852-01-15 7:00pm")
```


```{r parsed-solution}
dt1 <- "31/10/89"
dt2 <- c("2022-07-01 3:10am",
         "1852-01-15 7:00pm")
dmy(dt1)
ymd_hm(dt2)
```

## using `as_datetime`

Sometimes you can get away with calling as_datetime()

```{r}
as_datetime("2017-01-02 20:11:59")
```

But not always
```{r}
as_datetime("01-02-2017 20:11:59")
```

## using `strptime()` 

This is another convenient method to convert to date/time in baseR

```{r}
strptime("Monday, January 2,  2023", tz = "EST", format = "%A, %B %e, %Y")
```

## using `make_date` and `make_datetime`

These functions can make dates from individual (numerical) components:

```{r}
make_datetime(2024, 3, 5, 10, 00, 59, tz="EDT")
make_date(2024, 01, 10)
```

## The epoch

UNIX systems sometimes represent time as "the number of seconds which have elapsed since 1-1-1970." This date is known as "the epoch". So you may occasionally come across date-times that look like:

```{r}
now() |> as.integer()
```
To convert these to date format you can use as_datetime:

```{r}
as_datetime(3600 * 24)  # 24 hours after the epoch
```



## Parsing dates in files

We saw earlier the use of `cols(colname = col_date("EXPRESSION"))`. There is also `col_datetime` and `col_time`.

For these we need to give a pattern to match using various keys:

>* Year: four digit `"%Y"`, two digit `"%y"`
>* Month: number `"%m"`, abbreviation `%b`, full name `%B`
>* Day: number in month `"%d"`
>* Hours:minutes:seconds: `"%H:%M:%S"`
>* See `parse_datetime` for more

```{r}
parse_datetime("22/10/1, 10:01:00", 
               format = "%y/%m/%d, %H:%M:%S")
```

## Extracting components

```{r}
right_now <- now()
day(right_now) # same as mday()
month(right_now)
year(right_now)
yday(right_now) # day within year ("Julian date")
wday(right_now) # day within week, starting with Sunday
hour(right_now)
minute(right_now)
second(right_now)
```

## Time Math 

```{r}
ymd("2023/01/04") - mdy("12/16/2022")
```

The `seconds`, `minutes`, `hours`, `days`, `weeks` can be used to get offsets:
```{r}
right_now <- now()
right_now
right_now + minutes(10)
right_now + weeks(7)
```

## Time spans

After doing math, you might want to convert the units.

```{r}
# Days to Last day of class!
semester_end <- dmy_hms("22042024 23:59:59") - now()

(sed <- as.duration(semester_end))

as.integer(sed, "days")
as.numeric(sed, "minutes")

```

## Durations
There are equivalent `d*` functions (`dyears`, etc) for constructing durations directly.

```{r}
(d = today() - update(today(), year = 2023))
```
Anytime part can be converted to seconds using duration class

```{r}
as.duration(1234)
dseconds(10)
dweeks(3)
```


## Arithmetic with duration

Arithmetic with durations works as you would expect:

```{r}

2 * dyears(1)
dyears(1) + dweeks(12) + dhours(15)
```

You have to be careful when adding durations and dates. What is 1 day after next Saturday at 1pm?

```{r}
# this seems fine
next_saturday = mdy_hm("March. 8, 2024 13:00PM", tz="America/Detroit")
next_saturday + ddays(1)

# But what happened to this addition?
next_saturday = mdy_hm("March. 9, 2024 13:00PM", tz="America/Detroit")
next_saturday + ddays(1)
```

You can use methods of periods from lubridate instead

```{r}
next_saturday = mdy_hm("March. 9, 2024 13:00PM", tz="America/Detroit")
next_saturday + days(1)

# more significant with this example
ymd("2024-01-01") + years(1)
ymd("2024-01-01") + dyears(1)
```
## Exercise

Parse these dates.

* What's the range of the dates? 
* What is the mean date?
* What is the median date? 
* Suppose you want reminders of the anniversary of these dates in the coming year one month in advance. Find the dates of the reminders.

```{r diffex, exercise = TRUE}
ds <- c("1981-09-25", "1982-03-06", "2010-12-14", "2012-10-18", "2019-04-03")
```

```{r diffex-solution}
ds <- c("1981-09-25", "1982-03-06", "2010-12-14", "2012-10-18", "2019-04-03")

# parse as year-month-day format
ds_parsed <- ymd(ds)

# range
range(ds_parsed)
# or
min(ds_parsed)
max(ds_parsed)

# mean and median
mean(ds_parsed)
median(ds_parsed)

# setting reminders
## method 1
year_diff <- year(today()) - year(ds_parsed)
ds_this_year <- ds_parsed + years(year_diff) - months(1)
if_else(ds_this_year < today(), ds_this_year + years(1), ds_this_year) 

## method 2
ds_parsed_2 <- ds_parsed - months(1)
year(ds_parsed_2) <- year(today())
ds_parsed_2 + if_else(ds_parsed_2 < today(), years(1),  years(0))
```

## Revisiting of some common operations on Date columns

```{r}
aatemp <- read_csv("data/ann_arbor_weather.csv.gz")
class(aatemp$DATE)

## range of Dates in number of days
summarize(aatemp, as.numeric(max(DATE) - min(DATE), "days")) 

## Finding maximum TMAX per year-month 
aatemp |> group_by(year(DATE), month(DATE)) |> 
  summarize(m = max(TMAX, na.rm = TRUE)) |>
  arrange(desc(m)) |> head(5)

## limiting to a specific date range
aatemp |> filter(DATE >= "2018-01-01", DATE <= "2018-03-31") |>
  nrow() # 31 + 28 + 31
```

## Exercise

Using the `weather` table in the `nycfights13` package, compute the average temperature at each airport (`origin`) between 9am and 15pm on September 1, 2013 (see the `time_hour` column).

```{r nyc0901temp, exercise = TRUE}

```


```{r nyc0901temp-solution}

filter(weather, time_hour >= "2013-09-01 09:00:00", time_hour <= "2013-09-01 17:00:00") |>
  group_by(origin) |> summarize(mean(temp))
```


