---
subtitle: "Stats 306: Lecture 16"
title: "Strings: Regular Expressions"
author: "Jayashree Ravi"
citation: "Content adapted from slides by Dr. Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
set.seed(2939394)
if (!require(nycflights13)) install.packages("nyclfights13", repos = "http://cran.us.r-project.org")

sonnet18 <- "Shall I compare thee to a summer's day?\nThou art more lovely and more temperate:\nRough winds do shake the darling buds of May,\nAnd summer's lease hath all too short a date;\nSometime too hot the eye of heaven shines,\nAnd often is his gold complexion dimm'd;\nAnd every fair from fair sometime declines,\nBy chance or nature's changing course untrimm'd;\nBut thy eternal summer shall not fade,\nNor lose possession of that fair thou ow'st;\nNor shall death brag thou wander'st in his shade,\nWhen in eternal lines to time thou grow'st:\n\tSo long as men can breathe or eyes can see,\n\tSo long lives this, and this gives life to thee."

sonnet18_lines <- str_split(sonnet18, "\n")[[1]]
```

## Review: Relational Data

* Mutating joins: outer left, right, full and inner
* Keys spread across multiple columns (use `count(tbl, col1, col2)`; if n > 1 then `col1, col` do not form a primary key)
* "Natural joins": don't specify `join_by` argument, let R match column names (be careful!)
* Use anti-join and/or semi-join to use as filter

## Strings and Factors

R has two data types for representing categorical data types:

* Strings: character data
* Factors: integers representing categories with look up table for labels, can be ordered 

```{r}
example_str <- c("you", "say", "goodbye", "i", "say", "hello", "hello", "hello")
example_str
example_fct <- factor(example_str)
example_fct
```

## Strings vs. factors

Many times two types have the same function, why use one or the other?

- Want to order lexicographically: strings
- Want to order in other way: factors
- Treat as text (search, combine): strings
- Avoid errors from treating as text: factors

There used to be a space saving advantage to factors but that has been eliminated with newer versions of R.

## String basics

A string (or "character vector" in R) is character data enclosed in double `"` or single `'` quotes.

To include quotation marks inside, we need to **escape** them:
```{r}
baberuth <- "George Herman \"Babe\" Ruth Jr., a.k.a. \"the Bambino\" and \"the Sultan of Swat\""
baberuth
cat(baberuth) # prints directly instead of displaying as escaped string
```

## Other special characters

Since `\` is used to escape things, if you need a literal backslash use `\\`. 

>* New line: `\n`
>* Tab: `\t`
>* Unicode characters: `\unnnn` (u followed by 4 digit code)

This often come up in searching text for certain patterns.

## Combining strings

If we need to make strings out of other strings, we have two (more or less identical) functions:

* Old school: `paste` (joins with " ") and `paste0` (joins with "");
* New school: `str_c` (joins with ""); used in our book

```{r}
paste("Section", c("100", "200", "100"), 
      "meets on", 
      c("Tuesday", "Monday", "Wednesday"))
```
Notice: this is a vector of length 3, use `collapse = "SEP"` to make a single string from a vector.

## Exercise

Join the following using a new line characters `\n` (use the `sep` and `collapse` arguments as necessary). Demonstrate using `cat`.

```{r sonnet18, exercise = TRUE}
line1 <- "Nor shall death brag thou wander'st in his shade,"
line2 <- "When in eternal lines to time thou grow'st:"
line3 <- "So long as men can breathe or eyes can see,"
line4 <- "So long lives this, and this gives life to thee."
```

## Properties of strings

```{r, echo = FALSE}
line1 <- "Nor shall death brag thou wander'st in his shade,"
line2 <- "When in eternal lines to time thou grow'st:"
line3 <- "So long as men can breathe or eyes can see,"
line4 <- "So long lives this, and this gives life to thee."
```

The `length` function in R will tell you how many items are in a vector:

```{r}
sonnet18_end <- c(line1, line2, line3, line4)
length(sonnet18_end)
```

To find the length of the strings in terms of characters:
```{r}
str_length(sonnet18_end) # base::nchar also works
```

## Subsetting

The `str_sub` function can retrieve or change **substrings**.

```{r}
line1
str_sub(line1, 11, 20)
line2
str_sub(line2, 36)
str_sub(line2, -8)
line4
line4b <- line4
str_sub(line4b, 1, 10) <- "REMOVED"
line4b
```


## Splitting

Here's the full sonnet as a single string:
```{r}
sonnet18 <- "Shall I compare thee to a summer's day?\nThou art more lovely and more temperate:\nRough winds do shake the darling buds of May,\nAnd summer's lease hath all too short a date;\nSometime too hot the eye of heaven shines,\nAnd often is his gold complexion dimm'd;\nAnd every fair from fair sometime declines,\nBy chance or nature's changing course untrimm'd;\nBut thy eternal summer shall not fade,\nNor lose possession of that fair thou ow'st;\nNor shall death brag thou wander'st in his shade,\nWhen in eternal lines to time thou grow'st:\n\tSo long as men can breathe or eyes can see,\n\tSo long lives this, and this gives life to thee."
```

Let's split that up into individual lines as strings in a vector:
```{r}
sonnet18_lines <- str_split(sonnet18, "\n") # notice use of new line
sonnet18_lines
```

What is happening? Since `sonnet18` could have had multiple entries (vector), `str_split` returns a **list** with splits for each item in the vector:
```{r}
str_split(c("a b c", "d e"), " ")
```



## Exercise

**Tokenization** is the process many computer languages use to turn strings into discrete words. Use `str_split` to turn each line of `sonnet18_lines` into individual words. Use `unlist` on the result and answer the question: how many words are there?

```{r tokenize, exercise = TRUE}

```


## Looking for patterns: Regexes

A **REG**ular **EX**pression (**regex**) is a mini-language for writing patterns to match in strings.

They are used in R but also in other languages, the command line, databases, etc.

A regex is just a string, albeit one that looks rather strange.

## Regexes pattern matches

Alphanumeric characters (a-z, A-Z, 0-9) match directly, most other characters with some later exceptions.

```{r}
x = c("apple", "banana", "pear") %>% print
str_view(x, pattern = "an", html = TRUE)
```

`.` will match exactly one character (not newline)

```{r}
str_view(x, "a..", html = TRUE)
str_view("else every eele etcetera", "e..e") 
```

Limited choices with `[]`
```{r}
str_view(x, "a[pn]", html = TRUE)
```

Or classes of characters
```{r}
words
str_view(words, "a[b-d]", html = TRUE)
```

Negation choices with `[^]`
```{r}
str_view(words, "a[^b-d]", html = TRUE)
```

## Using str_extract

If we want to "extract" the first match we can use `str_extract()`

```{r}
str_extract("else every eele etcetera", "e..e ") 
```

To get all matches use str_extract_all()

```{r}
str_extract_all("else every eele etcetera", "e..e ") 
```


## Quantifiers

The four quantifiers you should know are:

>* ?: match zero or one of the preceding character.
>* +: match one or more of the preceding character.
>* *: match zero or more of the preceding character.
>* {x}: match exactly x of the preceding character.
>* {x,y}: match between x and y of the preceding character.
>* {x,}: match at least x of the preceding character.

```{r regex-ex1, exercise = T}
text = "Hello 200, this is a sample string with some random words and 123 numbers."
# Get all words that start with uppercase
# get all numbers
# get all words that are between 3-4 characters in length

```
## Character classes

A "character class" is a special pattern that matches a collection of characters.

>* \w: matches any alphanumeric character and is equivalent to using [A-Za-z0-9_]
>* \s: matches single space, tab, newline characters and is equivalent to using [\t\n\r\f\v]
>* \d: matches any digit and is equivalent to using [0-9]
>* \b: match a "word boundary" (more on this in a moment).

```{r regex-ex2, exercise = T}
str_view("this is a word", "\w")
```
Why did this throw an error?

```{r}
str_view("this is a word", "\\w")
 # OR
str_view("this is a word", r"(\w)")
```

Raw strings are preferred way of representing regex in almost all languages

```{r}
str_view(c("Rafael Nadal", "Roger Federer", "Novak Djokovic"), "\\b")
```

     
## Groups

To select between alternates use a vertical bar `|`. We often combine this with groups based on `()`:

```{r}

text = "It is raining cats and dogs"
str_view(text, "cat|dog", html = TRUE)
```

```{r}
text2 = "Get me a cat or get me a cap"
str_view_all(text2, "ca(t|p)", html = TRUE)
```

Repeat patterns in `( )` using `\\NUMBER` (**backreferences**). Contiguous words with same letter: 

```{r}
# get all those words that has a two letter pattern that it also ends with
str_view(words, "(..).*\\1$", html = TRUE)
```

## Anchors

Sometimes we want a match to occur at a particular position in the string. 

For example, "all words which start with b". 
For this we have the special anchor characters: ^ and $. The caret ^ matches the beginning of a string. The $ matches the end.

```{r}
x <- c('apple', 'banana', 'pear')
str_view(x, '^b')
str_view(x, 'r$')
```


## Exercise

Find all the words, in the `words` collection that start with consonant and end in a vowel

```{r eg2, exercise=TRUE}

```


## str_replace

`str_replace(v, re, rep)` will replace each match of re in v with rep. The most basic usage is as a sort of find and replace:

```{r}
str_replace('Give me liberty or give me death', '\w+$', 'pizza')
```


## Art and Emotion data set

```{r}
wikiart <- read_tsv("./data/WikiArt-Emotions-All.tsv.gz")
colnames(wikiart)
```

## Columns

With lots of columns, we could write things out direct, but better to use regex to find columns matching a pattern:

```{r}
select(wikiart, matches("anger"))
select(wikiart, matches("[tT]itle"))
```

## Making more tidy

This has lots of **ratings**, let's put them into better columns.

```{r}
pivot_longer(wikiart, 
             matches(":"), 
             names_to = c("rating_type", "emotion"),
             names_sep = ":",
             values_to = "mean_annotation") -> wa_long

select(wa_long, Title, rating_type, emotion, mean_annotation)
```

## Find and replace

Hard to see in previous but there are spaces in the emotions:

```{r}
wa_long$emotion[1:5]
```

Let's remove all white space
```{r}
wa_long <- mutate(wa_long,
                  mean_annotation = str_replace(emotion, "\\s", ""))
wa_long$emotion[1:5]
```







