---
title: "Stats 306: Lecture 8"
subtitle: "Wrapping up functions"
author: "Jayashree Ravi"
citation: "Content adapted from slides by Dr. Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this
library(Lahman) # install.packages("Lahman") if you don't have this
```

## Review

* From duplicated to clean code:
  * Common patterns in function
  * Use of `across`
  * Lists of functions as arguments
* Return values: only one but can package up in vectors, lists, and tables
* Predicates: functions that return `TRUE` or `FALSE`; use in `select`
* Anonymous functions, tidyverse short hand notation `~ (.x - mean(.x))^2 / (length(.x) - 1)`

## The `Lahman` package

Sean Lahman maintains an extremely [detailed database of American baseball
statistics](https://www.seanlahman.com/baseball-archive/statistics). The
`Lahman` package conveniently imports this data for us.

```{r}
library(Lahman)
glimpse(Batting)
dim(Batting)
colnames(Batting)
```

We can look up the description of the columns using `?Batting` or the help menu.

## Getting in the Lahman data

```{r}
Batting$yearID |> range()
group_by(Batting, playerID) |> 
    summarize(across(c("G", "HR", "RBI"), sum)) |>
    arrange(desc(HR)) |> head(5)
```

## From IDs to full names

The `People` table has a `playerID` column. Does this uniquely identify players?

```{r}
dim(People)
summarize(People, n_distinct(playerID))
```

Looking up players: 
```{r}
lookupPlayer <- function(pid) {
  filter(People, playerID %in% pid) # result will be a table
}

lookupPlayer(c("aaronha01", "ruthba01")) |> select(nameFirst, nameLast)
```

## `slice_max` function

Another way of slicing the top `n` values of any count is using `slice_max`

```{r}
Batting |> group_by(playerID) |> summarize(n = n()) |> slice_max(n = 3, n)
```

A few other variations are `slice_min`, `slice_sample`, etc. 

## Dataframe function caveats

A case of a problem of indirection, as dplyr uses tidy evaluation to allow you to refer to the names of variables inside your data frame without any special treatment.

```{r embracing, exercise = T}
grouped_mean <- function(df, group_var, mean_var) {
  df %>%
    group_by(group_var) %>%
    summarize(mean(mean_var))
}

grouped_mean(Batting, playerID, HR)
```

Fix with embracing `{{ }}`

```{r}
grouped_mean <- function(df, group_var, mean_var) {
  df |>
    group_by({{ group_var }}) |>
    summarize(mean({{ mean_var }}))
}

grouped_mean(Batting, playerID, HR) |> head()
```



## Successful players (batting)

Which players had the most home runs? Games played? At bats?

```{r}
top_players <- function(stat, type) {
  Batting |> group_by(playerID) |> 
    summarize(total = sum({{ stat }})) |> 
    slice_max(n = 1, total) |> 
    mutate(lookupPlayer(playerID)) |> 
    select(playerID, total, nameFirst, nameLast) |>
    mutate(type = type)
} 

# rbind stacks tables
rbind(
 top_players(HR, "Homeruns"),
  top_players(G, "Games"),
  top_players(AB, "At Bats"))

```




## Variable number of arguments

Recall our template for creating a function in R:

```{r, eval = FALSE}
fname <- function(required_arg, optiona_arg = 123, ...) {
  # function body
}
```

The `...` means "any number of additional arguments"

## Example of variable arguments

```{r}
sum_summary <- function(...) {
  select(...) |> summarize_all(sum)
}

sum_summary(Batting, HR, G, SB)
```

Notice how all three arguments get passed to `select` automatically.


## Exercise

The `paste` function will take more than one item to turn into a string using its `...` argument:

```{r}
paste("Welcome", "to", "Stats206", "and", "DataSci101")
```

Write a function that takes a `...` argument and passes it along to `paste`. Set the `sep = "+"` when calling `paste`.

```{r exdots, exercise = TRUE}


## plus_paste("Welcome", "to", "Stats306", "and", "R Programming")
```

## Common use of `...`: passing optional arguments to called functions

We have seen `na.rm = TRUE` pop up from time to time.

```{r}
d <- tibble(x = c(1, 10, NA, 7), y = c(1000, 2000, 10000, NA))
d
```

What happens if we try to summarize using `mean`?

```{r}
summarize_all(d, mean)
```
`summarize` uses `...` to pass additional arguments to the summarizing functions.

```{r}
summarize_all(d, mean, na.rm = TRUE)
```
## Fixing a previous function

```{r}
sum_summary <- function(...) {
  select(...) |> summarize_all(sum, na.rm = TRUE)
}

sum_summary(Batting, HR, G, SB)
```

## Conditional evaluation, short circuiting

We have used `if_else` to make choices for an entire vector. We often need **conditional program flow** in functions.

```{r}
my_median <- function(x) {
  x_sorted <- sort(x)
  n <- length(x)
  if (n %% 2 == 1) {
    ## odd numer of observations
    return(x_sorted[(n + 1)/2])
  } else {
    return(mean(x_sorted[c(n/2, n/2 + 1)]))
  }
  
  return(Inf) # this line will never run
}

values_even <- c(19, 2, 4, 11)
values_odd  <- c(19, 2, 4, 11, 9)

my_median(values_even)
my_median(values_odd)
```

## Careful about what you put in your `if` statements:

```{r eval = FALSE}

if (values_even > 3) {
  ... you may be trying to see if all values are larger than 3 ...
}
```
Would give the error message:
```
Error in if (values > 3) { : the condition has length > 1
```

Solution: use the `any` and `all` functions


## Combining more than one condition

We can use `&` for "and" and `|` for "or". We can use `!` to negate.

Here's an example where we can only set one option for a function:

```{r}
food_order <- function(entree, side = NULL, dessert = NULL) {
  if (!is.null(side) & !is.null(dessert)) {
    return("You may only order one side or dessert")
  } 
  
  extra <- 
    if (!is.null(side)) {
      paste("with a side of", side)
    } else if (!is.null(dessert)) {
      paste("with a dessert of", dessert)
    } else {
      # both NULL <=> nothing extra ordered
      "with no other items"
    }

  paste("You ordered a main dish of", entree, extra)
 
}

food_order("soup")
food_order("soup", side = "salad")
food_order("soup", dessert = "cake")
food_order("soup", "salad", "cake")
```

## Environments and functions

An **environment** in R is all defined variables and their values (we can think of it like a list). 

You have a **global** environment for your R session, and R packages and functions have their own environments.

```{r}
x <- 3
f <- function(y) {
  x <- 2
  y + x
}
f(2)

```

## Lexical scoping

Environments are **nested**. If we request a variable not in the current environment, we go up the chain. We call this lexical scoping.

```{r}

x <- 3
f <- function(y) {
  y + x
}
f(2)
```



## Exercise

Which columns contain a value greater than 100?

To find out, first, select only numeric columns (`is.numeric` will be helpful), then select the columns that have a value of at least 100 somewhere in column. Use `colnames()` to list the column names.

Then find the opposite. What columns do not have values greater than 100? 

```{r selectexercise, exercise = TRUE}

```

```{r selectexercise-solution}
select(Batting, where(\(x) is.numeric(x))) |> select(where(\(x) max(x, na.rm = TRUE) >= 100)) |> colnames()

select(Batting, where(~ is.numeric(.x))) |> select(where(~ max(.x, na.rm = TRUE) < 100)) |> colnames()

```
