---
title: "Stats 306: Lecture 3"
subtitle: "More data exploration with ggplot"
author: "Jayashree Ravi"
citation: Content adapted from slides by Dr. Mark Fredrickson
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
```

## Review: More on R

* Learned more about R's functions (required and optional arguments); `f(x, y =10)`
* Piping and assignment; `|>, =`
* Special values: `NA`, `NULL`, and `NaN`
* Vectors: ordered collections of same type, many computations are *vectorized*; See example below
* Lists: ordered collections of mixed types; `c(1:10); list(1, c(1:10))`
* Tables: lists of vectors of the same length (column oriented); `tibble, data.frame`
* Other items: indexing with `[]` and `$`, `dim` and `colnames` for tables

**Example On Vectorized operations**

Multiply all elements by 10

```{r vectorized, exercise = T}
v = c(1:10)
print(v)

```

## Graphing and Plotting

* Semiology of Graphics: Describes fundamental aspects of graphs, tasks
  * Visual variables (Position, Size, Value, Texture, Color, Orientation, Shape) 
  * Relationships/Tasks: Comparison, Correlation, Hierarchical, part-to-whole
* Grammar of Graphics, `ggplot2` library

```{r}
ggplot(data = mpg,
       aes(x = displ, y = hwy, color = cyl)) +
  geom_point()
```

**The components of a ggplot graph**

>* The graph object itself (creating using `ggplot(data)`)
>* A set of *aesthetic* mappings (connecting data to visual variables)
>* Layers: collections of geometric elements (`geom_*()`) and statistical transformations (`stat_*()`)
>* Scales: information on the range or composition of variables
>* Coordinate systems: how the data are arranged spatially
>* Facet: breaking a single plot into many, similar plots
>* Theme: all the other color and printing aspects of the plot

**Alternative forms**

```{r eval = F}
ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy, color = cyl))

mpg |> ggplot() + geom_point(aes(x = displ, y = hwy, color = cyl))

mpg |> ggplot(mapping = aes(x = displ, y = hwy, color = cyl)) + geom_point()

ggplot(mpg, aes(x = displ, y = hwy, color = cyl)) + geom_point()

ggplot(mpg, aes(x = displ)) + geom_point(aes(y = hwy, color = cyl))
```

Note: You can use positional arguments or use `data`, `mapping` argument names to call these functions

mapping values from `ggplot` are inherited by layers


## geom_point

When to use scatter plot?

Let's visit the docs: [`geom_point`](https://ggplot2.tidyverse.org/reference/geom_point.html)

## Trying out some other mappings

```{r}
ggplot(data = mpg,
       aes(x = displ, y = hwy,
           size = cyl,
           color = class)) + 
  geom_point()
```

## Other geometries

Points are great for graphing two **quantitative** or **continuous** variables.
For **discrete** or **qualitative** we need other tools.

```{r}
ggplot(data = mpg,
       aes(x = displ, y = class)) + 
  geom_boxplot() 
```

## More than one geometry

We can add more than one layer:

```{r}
ggplot(data = mpg,
       aes(x = displ, y = class)) + 
  geom_boxplot() +
  geom_jitter(aes(color = hwy), height = 0.1)
```
(NB: `height` is controlling amount of vertical jitter)

## List of geometries

[More ggplot documentation](https://ggplot2.tidyverse.org/reference/index.html#geoms).

## Statistical Summaries

* In addition to the raw data (or our calculations), our plots involved **data summaries** 
* `ggplot2` calls these **summary statistics** or `stat_*` functions
* We already saw a summaries in the boxplot: quantiles, twice IQR bars
* We can access summaries that geometries compute and add additional summaries.

## Boxplot as statistic

```{r}
ggplot(data = mpg,
       aes(x = displ, y = class)) + 
  stat_boxplot() 
```

Compare with

```{r}
ggplot(data = mpg,
       aes(x = displ, y = class)) + 
  geom_boxplot() 
```

## Adding computed summaries

The `stat_summary` function allows you to use any function to summarize 

```{r}
ggplot(data = mpg,
       aes(x = displ, y = class)) + 
  geom_boxplot() +
  stat_summary(fun = mean, size = 3, color = "red", geom = "point") 
```

## Trend lines

When using scatter plots, one of the most common summaries is a **trend line**.
```{r}
ggplot(data = mpg,
       aes(x = displ, y = hwy)) + 
  geom_point(position = "jitter", alpha = 0.25) +
  stat_smooth() # geom_smooth also works
```

## More layering

```{r}
ggplot(data = mpg, aes(x = displ)) +
  geom_point(aes(y = hwy), color = "orange") +
  geom_point(aes(y = cty), color = "blue") +
  stat_smooth(aes(y = hwy), lty = 1, color = "black") +
  stat_smooth(aes(y = cty), lty = 2, color = "red")
```

We'll see a better way to make this table when we talk about tall vs. wide format data.

## Overriding defaults of `stat` functions

Each `geom_*` has a default statistic function. We can override this.

```{r}
ggplot(data = mpg, aes(x = class)) +
  geom_bar() # default stat is count
```

## Average highway efficiency by class

```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  geom_bar(stat = "summary", fun = "mean") 
```

## Replacing tables

We often use **tables** in documents to give numerical summaries. But why not
replace those with a nice graphic?

```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  stat_summary(
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

## Exporting Plots

* Automatically included in knitted RMarkdown documents
* Export from the Plots panel in RStudio
* `ggsave` function to save to file (variety of formats)

