---
subtitle: "Stats 306: Lecture 15"
title: "Relational Data, Continued"
author: "Jayashree Ravi"
citation: "Content adapted from slides by Dr. Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
set.seed(2939394)
if (!require(nycflights13)) install.packages("nyclfights13", repos = "http://cran.us.r-project.org")

sonnet18 <- "Shall I compare thee to a summer's day?\nThou art more lovely and more temperate:\nRough winds do shake the darling buds of May,\nAnd summer's lease hath all too short a date;\nSometime too hot the eye of heaven shines,\nAnd often is his gold complexion dimm'd;\nAnd every fair from fair sometime declines,\nBy chance or nature's changing course untrimm'd;\nBut thy eternal summer shall not fade,\nNor lose possession of that fair thou ow'st;\nNor shall death brag thou wander'st in his shade,\nWhen in eternal lines to time thou grow'st:\n\tSo long as men can breathe or eyes can see,\n\tSo long lives this, and this gives life to thee."

sonnet18_lines <- str_split(sonnet18, "\n")[[1]]
```

## Review

* `nycflights` data: flights, airlines, airports, planes, weather
* Keys: **primary keys** identify data within the table, **foreign keys** link to data in other tables.
* Relations: tables created by **joining** two or more tables. Classify as: mutating or filtering.
* Types of joins: outer: left, right, and fulll; inner
* Use of `dplyr` `*_join` functions, `by` argument, `suffix` argument

## Example: What models of planes fly from NYC to Detroit?

```{r}
filter(flights, dest == "DTW") |>
  left_join(planes, by = "tailnum") |>
  count(paste(manufacturer, model)) |> arrange(desc(n))
```


## Inner joins

Inner joins discard entries where a key does not exist in the table. Recall not all destinations are in the airports table (international flights):

```{r}
fiveam <- filter(flights, year == 2013, month == 1, day == 1, hour == 5)
left_join(fiveam, airports, by = c("dest" = "faa")) |>
  select(flight, dest, name)
```

Inner join would discard that row:
```{r}
inner_join(fiveam, airports, by = c("dest" = "faa")) |>
  select(flight, dest, name)
```

## Exercise

Use an inner join to find out how many domestic flights there were during the period covered by the `flights` data set.

```{r innerjoin, exercise = TRUE}

```
```{r innerjoin-solution}
flights |> inner_join(airports, join_by(dest == faa)) |> nrow()
```


## Matching multiple keys

What would happen if the joining column `origin` is not included?

```{r}
left_join(fiveam, weather, by = c("year", "month", "day", "hour")) |>
  select(flight, origin.x, temp, wind_speed, origin.y)
```

In a left join, if the same key matches multiple tables in the right, the row gets duplicated for each.

Lesson: Be careful with natural joins and compound keys.


## Filtering Example
Using semi-join:

Suppose we want to find the top 10 destinations by number of flights:

```{r}
top_dest <- flights %>%
  count(dest, sort = TRUE) %>%
  head(10)
top_dest
```

If we *join* to the `top_dest` table, we can get flights to those destinations:

```{r}
left_join(flights, top_dest) |> # natural join on `dest` column
  select(flight, dest, n) -> flights_to_tops
flights_to_tops
```
Then filtering out the NAs
```{r}
filter(flights_to_tops, !is.na(n))
```

Do it one step with `semi_join`:

```{r}
semi_join(flights, top_dest) |>
  select(flight, dest)
```

## Exercise

Find the set of flights that were delayed by more than one hour (`dep_delay > 60`). Find the weather patterns associated with those flights and find if the wind_speed and delay are correlated

```{r naturaljoin, exercise = TRUE}
# you can use a natural join for flights and weather

```

```{r naturaljoin-solution}
flights |> filter(dep_delay > 60) |> inner_join(weather) |> select(dep_delay, wind_speed) |> drop_na() |> cor() 
```



## Relational data wrap up

* When data are stored in more than one table you need to **join**
* **Mutating joins** join tables to add columns (and some times rows!)
* **Filtering joins** use one table to select rows for another
* Outer left/right/full joins, inner joins
* Also useful for linking table to itself 
* Set operations less generally useful, but have some specific uses (e.g., vectors)

