---
title: "Stats 306: Lecture 7"
subtitle: "More on User Defined Functions"
author: "Jayashree Ravi"
citation: "Content adapted from slides by Dr. Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this
aatemp <- read_csv("data/ann_arbor_weather.csv.gz")
```

## Functions: mini-programs

Functions are miniature programs that take inputs and return a value.

* Reuse the same set of steps in different contexts (variables, data sets)
* Express key steps that should be done performed later (delayed computation)

## Avoding repetition: groups

If we want to repeat the same computation within groups we can avoid repetition (always our goal!) using `group_by`:

```{r}
aatemp |>
  group_by(year(DATE)) |>
  summarize(mean(TMAX - TMIN))
```

## Correlation

In the previous grouping example, we used the same variables in each group to make our computation. Sometimes we want to repeat the same computation on different variables.

One way of defining the *correlation coefficient* for two variables is:

$$r_{xy} = \frac{1}{n - 1} \sum_{i=1}^n \frac{x_i - \bar x}{s_x} \frac{y_i - \bar x}{s_y}$$
In other words, *the average product of Z-scores*

```{r}
aatemp |>
  mutate(zx = (TMAX - mean(TMAX)) / sd(TMAX),
         zy = (TMIN - mean(TMIN)) / sd(TMIN),
         zz = zx * zy) |>
  summarize(sum(zz) / (length(zz) - 1))

# validate
summarize(aatemp, cor(TMAX, TMIN))
```

But lots of duplication!

## Writing our own functions

```{r eval = F}
function_name <- function(required_argument, optional_argument = 0, ...) {
  # function body
  return(a_value)
}
```

(NB: The last line is automatically returned.)

## Exercise

Write an R function to transform a vector into a $Z$-score
```{r scale, exercise = TRUE}
x <- c(-1, 2, 1.1)
```

```{r scale-solution}
z_score <- function(input) {
  (input - mean(input)) / sd(input)
}

z_score(x)
```


(NB: in the last lecture we used the `scale` function instead of computing this ourselves)

## Revisiting correlation:

```{r}

z_score <- function(input) {
  (input - mean(input)) / sd(input)
}

aatemp |>
  mutate(zx = z_score(TMAX),
         zy = z_score(TMIN),
         zz = zx * zy) |>
  summarize(sum(zz) / (length(zz) - 1))
```


## Common use case: mutations and summaries

Suppose we want to median center and scale in terms of the IQR. We are going to use this functionality on the original data, but also within years.

```{r}
myscale <- function(x) {
  (x - median(x)) / IQR(x)
}
```
```{r}
mutate(aatemp, TMAX_myscale = myscale(TMAX)) |>
  ggplot(aes(x = factor(year(DATE)), y = TMAX_myscale)) +
    geom_violin()
```


```{r}
aatemp |> group_by(year(DATE)) |>
  mutate(TMAX_scaled = myscale(TMAX)) |>
  ggplot(aes(x = factor(year(DATE)), y = TMAX_scaled))+
    geom_violin()
```



## Exericse: Converting all Fahrenheit to Celsius

We have both `TMAX` and `TMIN` in Fahrenheit.

```{r}
select(aatemp, TMAX, TMIN)
```

Let's convert to Celsius ($C = (F - 32) \cdot 5/9$) by first creating a function.

```{r celsius, exercise = TRUE}

```

## From duplicated code to clean code

How many interquartile ranges above or below the median? (Similar to a Z-score)

```{r}
mutate(aatemp, TMAX_scale_IQR = (TMAX - median(TMAX)) / IQR(TMAX),
               TMIN_scale_IQR = (TMIN - median(TMIN)) / IQR(TMIN)) |>
  group_by(abs(TMAX_scale_IQR) > 1.5, abs(TMIN_scale_IQR) > 1.5) |>
  summarize(n())
```

## Writing our own functions: capturing common behavior

```{r}
scale_IQR <- function(x) {
  (x - median(x)) / IQR(x)
}

temp_summary <- function(tbl) {
  group_by(tbl, abs(TMAX_scale_IQR) > 1.5, abs(TMIN_scale_IQR) > 1.5) |>
  summarize(n())
}

mutate(aatemp, TMAX_scale_IQR = scale_IQR(TMAX), 
               TMIN_scale_IQR = scale_IQR(TMIN)) |>
  temp_summary()
```

## Even better: `across`

We're still calling the function twice. `across` lets us pass in a
**function** as an argument!

If we give one function, it uses the original column names:

```{r}
mutate(aatemp, across(c("TMAX", "TMIN"), scale_IQR)) |>
    group_by(abs(TMAX) > 1.5, abs(TMIN) > 1.5) |>
    summarize(n())
```

## More than one function, giving them names

We can give a list of more than one function of the form
```{r eval = FALSE}
list(name1 = function1, name2 = function2)
```

```{r}
mutate(aatemp, across(c("TMAX", "TMIN"), 
          list(scale_IQR = scale_IQR, scale = scale))) |>
  colnames()
            
```

## Exercise

**Skew** measures the relative amount of variation above and below the center of distribution. The sample **coefficient of skew** is given by 

$$ \frac{(1/n) \sum_{i=1}^n (X_i - \bar X_n)^3}{\hat \sigma^3}$$
where $\hat \sigma$ is the sample standard deviation.

Write a function to compute the coefficient of skew and use `across` to compute it for `TMAX` and `TMIN` in the `aatemp` data set.

```{r coefskew, exercise = TRUE}

```

```{r coefskew-solution}
skew <- function(x) {
   mean((x - mean(x))^3) / sd(x)^3
}


summarize(aatemp, across(c("TMAX", "TMIN"), skew))
```

## Return values

R functions can return (at most) one object. Can we return several items? Yes: using vectors, lists, and tables.

```{r}
first_last <- function(x) { c(x[1], x[length(x)]) }
first_last(LETTERS)

mean_range <- function(x) { list(mean = mean(x), range = range(x))}
mr <- mean_range(rnorm(100)) # 100 random values
mr$mean
mr$range

numbered_table <- function(x) { 
  tibble(idx = seq_along(x), main_column = x)
}
numbered_table(letters)
```


## Summarizing with functions that return more than one value

```{r}
mean_var <- function(x) {
  tibble(average = mean(x), variance = var(x))
}
aat_year <- group_by(aatemp, year(DATE))
summarize(aat_year, mean_var(TMAX))
```

Use caution when using this function more than once! Now what are these values we got??

```{r}
summarize(aat_year, mean_var(TMAX), mean_var(TMIN))
```

## Predicates

A **predicate** is a function that returns `TRUE` or `FALSE`.

For example:
```{r}
no_missing <- function(x) {
  !(any(is.na(x)))
}

no_missing(c(2,3,44))
no_missing(c(7, NA, 122))
```

## Selecting columns with predicates

We have seen several predicates used already: `is.numeric` or `is.character` in selecting columns. We can also use our own predicates.

```{r}
select(aatemp, where(no_missing)) |> colnames()
```

## More on anonymous functions

Sometimes we want to write short little functions that only get used once. We call these **anonymous functions** like this:

```{r}
select(aatemp, where(function(x) { !any(is.na(x)) })) |> ncol()
```

## Tidyverse anonymous function short hand

Many of the tidyverse functions support using formulas to indicate an anonymous function:

```{r}
select(aatemp, where(~ !any(is.na(.x)))) |> ncol()
```

>* `~` starts a formula
>* `.x` is the argument to the function

Another form using backslash

```{r}
select(aatemp, where(\(x) !any(is.na(x)))) |> ncol()
```

We'll see more of this form when we discuss iteration.

**More function short hand**

```{r}
summarize_at(aatemp, c("TMAX", "TMIN"), ~ mean((.x - mean(.x))^3) / sd(.x)^3)
```

## Functions with optional arguments

Recall that R functions with `argument = value` are optional. We can write functions this way too.

```{r}

multSomething <- function(x, to_mult = 1) {
  x * to_mult
}

multSomething(10)
multSomething(10, 2)
multSomething(to_mult = 2, 10)

mutate(aatemp, TMAX1 = multSomething(TMAX),
               TMAX2 = multSomething(TMAX, 2)) |> 
  select(TMAX, TMAX1, TMAX2)
```

## Some functions allow passing additional arguments indirectly

```{r}
mutate(aatemp, across(c("TMAX", "TMIN"), \(x) multSomething(x, to_mult = 0))) |>
  select(TMAX, TMIN)
```



## Exercise

Write a function `center` that **takes a function** as an argument computes
$$x - f(x)$$
Make the default argument be `mean`, but demonstrate using `median` centering as well.

```{r higher-order, exercise = TRUE}
myvec <- c(203404, 292, 1010, 3, -10930, 39)
```

```{r higher-order-soultion}
myvec <- c(203404, 292, 1010, 3, -10930, 39)

center <- function(x, f = mean) {
  x - f(x)
}

center(myvec)
center(myvec, mean)
center(myvec, f = median)
```
